---
title: "Lesson 1 Basics"
author: "Julie Raiguel"
date: "January 16, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## R Markdown

"tgpp" is the variable used for the dataset.  The CSV file was read from the data directory within the project directory.  When I ran this during class with only one dot it worked, but when I opened it at home, it would only work when I added a second dot.  To make this code useful from other machines, I decided to read the dataset using the URL instead.
```{r}
tgpp <- read.csv("https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/tgpp.csv")
```


The "names" function returns the headers.
```{r}
names(tgpp)
```


The "nrow" and "ncol" functions return the numbers of rows and columns.
```{r}
nrow(tgpp)
ncol(tgpp)
```


The "apply" functions execute a function for each of the columns in the dataset.  In this case, I needed to know the classification of the data in the column.  "lapply" returns the output as a list, which I find easier to read than the simplified version outputted by "sapply".
```{r}
lapply(tgpp, class)
sapply(tgpp, class)
```


This code tells the computer to read the cells in row 1 and column 3, row 5 and column 7, and row 8 and column 10.
```{r}
tgpp[c(1,5,8), c(3,7,10)]
```


The "pdf" command should make a pdf of the plot with the name "richness_scale_plot1.pdf" in the working directory.

The next line plots the data from the "Scale" column on the x-axis and the "Richness" column on the y-axis.  The x and y axes are labeled the same as the respective columns, but the units of "Scale" are indicated.  Only points are included, not lines, and the color of these points is "seagreen4".
```{r}
pdf('./richness_scale_plot1.pdf')
plot(tgpp$scale, tgpp$richness, xlab = 'Scale (m^2)', ylab = 'Richness', type = 'p', col = 'seagreen4')
```


I repeat the plot line below so that the output is visible in the html file, not just in the new pdf file.
```{r}
plot(tgpp$scale, tgpp$richness, xlab = 'Scale (m^2)', ylab = 'Richness', type = 'p', col = 'seagreen4')
```


This plot is the same as the one above, except for the color of the points and the argument "log= 'xy'".  This argument changes the plot so both the x and y coordinates are plotted on a logarithmic scale. 

It is now not so skewed due to the wide range of areas that were sampled, so trends in the data are easier to see in the plot.  R omits the four data points where the richness equals 0 because these are undefined when converted to log scale.
```{r}
plot(tgpp$scale, tgpp$richness, xlab = 'Scale', ylab = 'Richness', type = 'p', col = 'springgreen2', log='xy')
```
