---
title: "Lesson 1 Intermediate"
author: "Julie Raiguel"
date: "January 17, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
***
  
###Question 1  
output is a matrix where each row is for a unique species, and the value in each column of that row is the average of that column for that particular species in the original dataset.
  
***
  
###Question 2

```{}
Set sp_ids equal to the unique values in the column Species of the iris dataset

Set output as a matrix where all values are 0, the number of rows is the same as the number of unique 
values identified in sp_ids

Set the names of the rows as each of the unique values identified in sp_ids
Set the names of the columns as the names of the columns in iris except the last one

Set i equal to 1 through the number of unique species in sp_ids
    Set iris_sp as a subset of iris where the rows are only those of the particular value of the sp_ids 
    indicated by i's value at this time and the columns are all of those in iris except Species
    Set j as 1 through the number of columns in the subset
        Set the variable x equal to zero
        Set the variable y equal to zero
        If the number of rows in the subset is greater than zero
            Set k equal to 1 through the number of rows in the subset
                Set the variable x equal to the sum of x's current value and the value of the cell in the 
                subset at the row k and column j
                Set the variable y equal to y's current value plus one
                Repeat for all values of k (through all of the rows in the subset)
            Set the value in the output matrix at row i and column j equal to x divided by y
            Repeat for all values of j (all columns in the iris_sp subset and the output matrix)
    Repeat for all values of i (all of the species in sp_ids and rows in the output matrix)
Print the output matrix

```
  
***
  
###Question 3   
* output could be renamed Species_Averages  
* x could be renamed Species_Sum  
* y could be renamed Total_Count  
  
***
  
###Question 4  
The final for loop can be avoided by using the mean function instead of keeping a sum and counter.
```{r}
sp_ids = unique(iris$Species)

output = matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(output) = sp_ids
colnames(output) = names(iris[ , -ncol(iris)])

for(i in seq_along(sp_ids)) {
    iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species)
    for(j in 1:(ncol(iris_sp))) {
        if (nrow(iris_sp) > 0) {
           output[i, j] = mean(iris_sp[,j])
        }
    }
}

output

```
  
***
  
###Question 5  
This for loop only works if x is a vector of integers with a length of 10 indices.
```{r}
x = c(1:10)
y = vector("integer", 10)
z = 0

for(i in 1:10) {
    z = z + x[i]
    y[i] = z
}

print(y)
```
  
***
  
###Question 6  
Now, when the sum is higher than 10, y is printed as NA at those indices.
```{r}
x = c(1:10)
y = vector("integer", 10)
z = 0

for(i in 1:10) {
      z = z + x[i]
      if(z>10) {
          y[i] = NA
      }
      else {
          y[i] = z
      }
}

print(y)
```
  
  
***
  
###Question 7  
This function works regardless of the length of x, as long as the values in the vector are integers.  If x is set as a matrix or vector of characters, an error message is printed for the user.
```{r, error = TRUE}
Vector_Sum = function(My_Vector) {
      if (class(My_Vector) %in% ('integer')){
          y = NULL
          z = 0
          for(i in seq_along(My_Vector)) {
              z = z + My_Vector[i]
              y[i] = z
          }
      return(y)
      }
      else {
          stop('My_Vector must be an integer vector')
      }
}

x = c(1:22)
Vector_Sum(x)

x=c(6:18)
Vector_Sum(x)

x = matrix(0, nrow =4, ncol = 6)
Vector_Sum(x)

x = c("a", "b", "c")
Vector_Sum(x)
```
  
***
  
